Fractional Types

### 2019-11-21 fracGC/p.tex
before merge
sec 6.1

  We propose a novel approach based on the idea of \emph{fractional
    types}. As a simple intuitive example, allocation of a new boolean
  value initialized to \textsf{false} also creates a value
  $\oneover{\textsf{false}}$ that can be thought of as a garbage
  collection (gc) process specialized to reclaim, and only reclaim,
  storage containing the value $\textsf{false}$. This gc process is a
  first-class entity that can sliced and diced by combining it with
  other gc processes and decomposing it in smaller
  processes.

@@

\noindent The circuit initializes the ancilla to \textsf{true} and
checks that it remains \textsf{true} after interacting with the two
inputs. If either one of the inputs is \textsf{true} (but not both)
the ancilla bit will be negated and the runtime check would fail. Thus
the circuit can be used as a runtime assertion that the two inputs are
equal. When applied to quantum inputs, the circuit detects whether the
two qubits are entangled or not. By varying the initial value of the
ancilla, it is possible to customize the dynamic assertion to check
for particular entangled states~\cite{DBLP:journals/cal/ZhouB19}.

@@

To make the space usage explicit in every step of evaluation, we
define an abstract machine whose states are pairs of a code register
and a memory register. In a low-level implementation the code register
would contain a pointer to the current instruction and the memory
register would contain a bit-level runtime representation of values
with one bit for boolean values, etc. In the Agda formalization, we
include the current combinator in the code register, and represent the
memory register using an index into an enumeration. Thus a memory
register containing a boolean value would be represented in Agda as 𝔽
::  𝕋 :: [] [ Fin.0F ] or 𝔽 :: 𝕋 :: [] [ Fin.1F ] depending on whether
the values is \textsf{false} or \textsf{true}. The prepending list
contains all possible values of the type and the number in between [\ldots ] 
is an index into the list, representing a finite number that
is guaranteed to be in bounds. In a low-level runtime representation
only a binary encoding of this finite would need to be stored but in
order to just prove that space is preserved, we can just inspect the
size of the prepended list. Putting it all together, machine states and
small-step reductions are formalized as follows:

@@

By lifting the scoping restriction, the development in the previous
section is already more general than the state of the art in ancilla
management.  It however shares the same limitation of requiring a
runtime check to ensure ancillae values are properly restored to their
allocation~\cite{10.1007/978-3-319-20860-2_13,Green:2013:QSQ:2491956.2462177}.
In this section, we address this limitation using a combination of
pointed types, singleton types, monads, and comonads.

@@

@@

Once singleton types are built, we need to make sure circuit
evaluation is also reflected in the type system. So in fact
computation itself is reflected in the type system. In order to mix
singleton types with other types, we use pointed types as a unifying
substrate. Now all types have values; we can focus (using monad) and
unfocus using (comonad). (This by the way ensures we don't write 1/0
which woudl be bad).

@@

>>>>

  We propose a novel approach based on the idea of \emph{fractional
    types}. As a simple intuitive example, allocation of a new boolean
  value initialized to \AgdaInductiveConstructor{false} also creates a value
  $\oneover{\AgdaInductiveConstructor{false}}$ that can be thought of as a garbage
  collection (gc) process specialized to reclaim, and only reclaim,
  storage containing the value $\AgdaInductiveConstructor{false}$. This gc process is a
  first-class entity that can sliced and diced by combining it with
  other gc processes and decomposing it in smaller
  processes.

@@

\noindent The circuit initializes the ancilla to \AgdaInductiveConstructor{true} and
checks that it remains \AgdaInductiveConstructor{true} after interacting with the two
inputs. If either one of the inputs is \AgdaInductiveConstructor{true} (but not both)
the ancilla bit will be negated and the runtime check would fail. Thus
the circuit can be used as a runtime assertion that the two inputs are
equal. When applied to quantum inputs, the circuit detects whether the
two qubits are entangled or not. By varying the initial value of the
ancilla, it is possible to customize the dynamic assertion to check
for particular entangled states~\cite{DBLP:journals/cal/ZhouB19}.

@@

To make the space usage explicit in every step of evaluation, we
define an abstract machine whose states are pairs of a code register
and a memory register. In a low-level implementation the code register
would contain a pointer to the current instruction and the memory
register would contain a bit-level runtime representation of values
with one bit for boolean values, etc. In the Agda formalization, we
include the current combinator in the code register, and represent the
memory register using an index into an enumeration. Thus a memory
register containing a boolean value would be represented in Agda as 𝔽
::  𝕋 :: [] [ Fin.0F ] or 𝔽 :: 𝕋 :: [] [ Fin.1F ] depending on whether
the values is \AgdaInductiveConstructor{false} or \AgdaInductiveConstructor{true}. The prepending list
contains all possible values of the type and the number in between [\ldots ] 
is an index into the list, representing a finite number that
is guaranteed to be in bounds. In a low-level runtime representation
only a binary encoding of this finite would need to be stored but in
order to just prove that space is preserved, we can just inspect the
size of the prepended list. Putting it all together, machine states and
small-step reductions are formalized as follows:

@@

By lifting the scoping restriction, the development in the previous
sections is already more general than the state of the art in ancilla
management.  It however shares the same limitation of requiring a
runtime check to ensure ancillae values are properly restored to their
allocation~\cite{10.1007/978-3-319-20860-2_13,Green:2013:QSQ:2491956.2462177}.
In this section, we address this limitation using a combination of
pointed types, singleton types, monads, and comonads.

@@

The toy language has two types (natural numbers and booleans) and two
functions (and their compositions). Say we wanted to prove that
\AgdaInductiveConstructor{compose isZero square} always returns
\AgdaInductiveConstructor{false} when applies to a non-zero natural
number. We can certainly do this proof in Agda (i.e., in the
meta-language of our formalization) but we would like to do the proof
within the framework of the toy language itself. The most important
reason is that this proof can be now be used within the language to
optimize programs (or as the case will be for the full language to
remove a runtime check).

The strategy we adopt is to create a lifted version of the toy
language with \emph{pointed types}~\cite{XXX}, i.e., types paired with
a value of the type. In such the lifted language, the evaluation
function has an interesting type: it keeps track of the result of
evaluation within the type:

\Jexamplecont{}

Now various properties of our small example
\AgdaInductiveConstructor{compose isZero square} can be derived within
the extended type system for the toy language:

\Jexampletest{}

Evidently we can track concrete values as the first two tests
show. What is more interesting is \AgdaFunction{test3} which shows a
property that holds for all natural numbers $n$. The full evaluation
was not used but only partially as needed to derive the property. This
form of partial evaluation will prove to be quite expressive allowing
us to keep track of ancillae values throughout complex programs as
demonstrated in Sec.~\ref{sec:cat}.

@@

Once singleton types are built, we need to make sure circuit
evaluation is also reflected in the type system. So in fact
computation itself is reflected in the type system. In order to mix
singleton types with other types, we use pointed types as a unifying
substrate. Now all types have values; we can focus (using monad) and
unfocus using (comonad). (This by the way ensures we don't write 1/0
which would be bad).

@@

revrev: rematerialization ! gc the gc process!
